Concurrent Programming with CXL

### 2020-06-06 book/paper.tex
Added a dining philosopers chapter

\begin{quote}
\begin{itemize}
\item Is Peterson's Algorithm blocking or busy-waiting?
\item Check your answer by developing a test.
\end{itemize}
\end{quote}

@@

@@

It would be easy to solve using a busy waiting
approach, but again, we would like to avoid busy waiting.
Introduced by the famous Dutch computer scientist Edsger Dijkstra,
a \emph{semaphore} is a primitive that fits the bill.  A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
units of a resource initially available.
When allocating a resource, a process decrements the
counter using the \texttt{procure} or simply \texttt{P} operation.  The
\texttt{procure} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{vacate} or {V} operation.
Figure~\ref{fig:semaphore} shows the \texttt{synch} module implementation of
semaphores.

>>>>

@@

\chapter{Deadlock}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../Diners.cxl}
\end{code}
\caption{Dining Philosophers.}
\label{fig:diners}
\end{figure}

When multiple processes are synchronizing access to shared resources, they
may end up in a \emph{deadlock} situation where one or more of the processes
end up being blocked indefinitely because each is waiting for another to give
up a resource.
The famous Dutch computer scientist Edsger W.~Dijkstra illustrated this using
a scenario he called ``Dining Philosophers.''

Imagine five philosopers sitting around a table, each with a plate of food in
front of them, and a fork between each two plates.  Each philosopher requires
two forks to eat.  To start eating, a philosopher first picks up the fork on
the left, then the fork on the right.  Each philospher likes to take breaks
from eating to think for a while.  To do so, the philosopher puts down both
forks.  The philosophers repeat this procedure.  Dijkstra had them repeating
this for ever, but for the purposes of this book, philosphers can leave
the table when they're not eating.

Figure~\ref{fig:diners} implements the dining philosophers in CXL, using a
process for each philosopher and a lock for each fork.  If you
run it, CXL complains that the execution may not terminate, with all five
processes being blocked trying to acquire the lock.

\begin{quote}
\begin{itemize}
\item Do you see what the problem is?
\item Does it depend on \texttt{N}, the number of philosophers?
\item Can you come up with a solution?
\end{itemize}
\end{quote}

We will discuss ways of preventing deadlock later, but one important
approach to avoiding deadlock is to establish an ordering among the
resources (in this case the forks) and, when needing more than one
resource, always acquiring them in order.  In the case of the philosopers,
they could prevent deadlock by always picking up the lower numbered fork
before the higher numbered fork, like so:

\begin{code}
\begin{verbatim}
    if left < right:
        call lock &(forks[left]);
        call lock &(forks[right]);
    else:
        call unlock &(forks[right]);
        call unlock &(forks[left]);
    ;
\end{verbatim}
\end{code}

\begin{quote}
\begin{itemize}
\item Try it out!
\end{itemize}
\end{quote}

@@

It would be easy to solve using a busy waiting
approach, but again, we would like to avoid busy waiting.
Introduced by the Dijkstra,
a \emph{semaphore} is a primitive that fits the bill.  A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
units of a resource initially available.
When allocating a resource, a process decrements the
counter using the \texttt{procure} or simply \texttt{P} operation.  The
\texttt{procure} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{vacate} or {V} operation.
Figure~\ref{fig:semaphore} shows the \texttt{synch} module implementation of
semaphores.
